---
title: "DSCaseStudy-1"
author: "Rajni Goyal"
date: "October 28, 2016"
output: 
    html_document:
      keep_md: true
---


# Introduction

####Our partners at "The World Bank" have captured and provided us with GDP Ranking and Education statistics of countries across the world.Both are independent data sets providing different types of data for more or less the same countries (How closely these data sets match up will be discussed further in this study). Although we have been provided the data from "The World Bank", there are several steps needed to be taken to tidy up the data before analysis can begin on the data. To ensure that the data is used in the most productive and meaningful manner that can increase the intrinsic value the special importance have been given to data quality. In the below sections, we will walk through loading and cleaning the datasets provided to us. Once we are sure that the data is tidy enough, we will dive into our analysis and answer 5 core questions.

____________________________________________

##Questions to be Analyzed

#### 1. Match the data based on the country shortcode. How many of the IDs match?
#### 2. Sort the data frame in ascending order by GDP (so United States is last). What is the 13th country in the          resulting data frame?
#### 3. What are the average GDP rankings for the "High income: OECD" and "High income: nonOECD" groups?
#### 4. Plot the GDP for all of the countries. Use ggplot2 to color your plot by Income.Group.
#### 5. Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group. How many countries are      lower middle income but among the 38 nations with the highest GDP?

<br>

## Required Libraries

#### The "downloader", "dplyr"", "ggplot2"", and "knitr"" libraries are required to carry out this study.If any of these packages have not been installed, install.packages lines below needs to be uncommented before knitting this file.

    #install.packages("downloader")
    #install.packages("knitr")
    #install.packages("ggplot2")
    #install.packages("dplyr")

    library(downloader)
    library(ggplot2)
    library(dplyr)
    library(knitr)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(downloader)
library(dplyr)
library(ggplot2)
library(knitr)
```

# Methods

____________________________________________________________

### GDP Data: Download and tidy up the data

```{r}
source('C:/DSCaseStudy-1Project/Analysis/Data/GDPData.R', echo = TRUE)
```
### Income Data: Download and tidy up the data

```{r}
source('C:/DSCaseStudy-1Project/Analysis/Data/IncomeData.R', echo = TRUE)
```


# Results

____________________________

### Merging of the GDP and the Income Data sets

```{r}
source('C:/DSCaseStudy-1Project/Analysis/Data/MergeData.R', echo = TRUE)
```

#### Question-1 : Merge the data based on the country shortcode. How many of the IDs match?

```{r}
## Indicate how many of the IDs match
length(intersect(GDPdata$Country.ShortCode, Income1$Country.ShortCode))
## Indicate how many of the rows contain NAs
sum(!complete.cases(MergeData))

## Remove rows with missing data
MergeData1 <- MergeData[complete.cases(MergeData),]
# Provide row count after removing rows with missing data
nrow(MergeData1) 
```

#### Answer-1 : Before removing the NA values, there are 224 matching IDs. Once all 49 NAs are removed, however, there remain 189 matching country code IDs.

___________________________________________________

#### Question-2 : Sort the data frame in ascending order by GDP (so United States is last). What is the 13th country in the resulting data frame?

```{r}
#### To sort the data in ascending order by GDP, the *order* function is used.
MergeData1 <- MergeData1[order(MergeData1$GDP.Millions.of.US.Dollars, decreasing = FALSE),] # Sort the data frame by GDP
MergeData1$Short.Name[13] # Display only the 13th country in the data frame
```

#### Answer-2 : After sorting the data in ascending order by GDP, St. Kitts and Nevis is the 13th country in the data frame.

___________________________________________________________

#### Question-3 : What are the average GDP rankings for the "High income: OECD" and "High income:nonOECD" groups?

```{r}
#### The *mean* function is used to calculate the average rankings for the *High income: OECD* and *High income: nonOECD* groups.
mean(subset(MergeData1, Income.Group == "High income: OECD")$Country.Rank)    # Calculate High income: OECD mean country rank
mean(subset(MergeData1, Income.Group == "High income: nonOECD")$Country.Rank) # Calculate High income: nonOECD mean country rank
```

#### Answer-3 : The average "High income: OECD"" GDP ranking is 32.96667 and the average "High income: nonOECD" GDP ranking is 91.91304.

______________________________________________________________

#### Question-4: Plot the GDP for all of the countries. Use ggplot2 to color your plot by Income Group.

```{r}

#### The "ggplot" function is used to render side-by-side boxplots of all countries' GDP by "Income.Group".
ggplot(data = MergeData1, aes(x=Income.Group, y=GDP.Millions.of.US.Dollars, fill=Income.Group)) +   geom_boxplot() +  
  theme(axis.text.x = element_text(angle = 55, hjust = 1, size = 12)) +                       
  xlab("Income Group") + ylab("GDP (Millions of US Dollars)") + ggtitle("GDP for All Countries by Income Group") 
# fill option fills the color by Income.Group, geom_boxplot creates boxplots, theme(axis) adjusts x axis label size and position, xlab and ylab provide labels and ggtitle provides the label to the main plot
```


#### The first boxplot visualization depicts all "GDP.Millions.of.US.Dollars"" data by "Income.Group". However, the data appears heavily right-skewed and large outliers in "High income: OECD"" and "Lower middle income" make it difficult to compare each distribution by "Income.Group".

#### For this reason, the second visualization of boxplots is rendered below to show the "GDP.Millions.of.US.Dollars" data after logarithmic transformation. By applying a natural log transformation to the data, the effects of heavy skewness and outliers are removed. This provides a clearer comparison of distributions from one income group to another. The horizontal line through each group's box represents the group's median and the red diamond represents the group's mean. Finally, the individual log transformed GDP values are added, with jitter, to further aid in reviewing differences in spread.

```{r}
## Render ggplot2 boxplot for log-transformed GDP data
ggplot(data = MergeData1, aes(x=Income.Group, y=log(GDP.Millions.of.US.Dollars), fill=Income.Group)) + 
  geom_boxplot() + stat_summary(fun.y=mean, geom="point", shape=23, size=3, fill="red") +    
  geom_jitter(shape=16, position=position_jitter(0.3)) +                                    
  theme(axis.text.x = element_text(angle = 55, hjust = 1, size = 12)) +                      
  xlab("Income Group") + ylab("Log Transformed GDP (Millions of US Dollars)") + ggtitle("Log Transformed GDP for All Countries by Income Group") 

```


#### Answer-4 : In examining the log-transformed visualization, it is clear most of the "High income: OECD"" group data exceed the remaining groups' data in terms of GDP since "High income: OECD's"" first quartile (Q1) location appears to be greater in GDP value than all other groups' Q3 locations (Though "High income: OECD's"" Q1 appears to be nearly identical to "Upper middle income's"" Q3). That is, 75% of all "High income: OECD"" data is greater than at least 75% of all other groups' data.

_____________________________________________________

#### Question-5: Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group.How many countries are Lower middle income but among the 38 nations with highest GDP?

```{r}
## Create additional data frame and add quantile column
MergeData2 <- MergeData1
MergeData2$GDP.Quantile <- ntile(MergeData2$Country.Rank, 5) 
MergeData2$GDP.Quantile 



## Generate table by Income.Group output
table(MergeData2$Income.Group, MergeData2$GDP.Quantile, dnn = c("Income.Group","GDP.Quantile"))
```


#### Answer-5 : Based on the newly added quantile groupings, 5 countries from the "Lower middle income" group are among the 38 nations with the highest GDP (quantile 1).

# Conclusion

___________________________________________________________________________

#### Important information is being provided by GDP and Income data sets regarding numerous countries throughout the world. By combining these data sets, powerful insights are gained into the relationship between each country's income group classification and GDP ranking. When analyzing these data sets, it becomes apparant that both data sets do not align perfectly in regards to country codes and relevant data. With the Cleaned Data, we answered some core questions from. We were provided with 189 countries containing data in both GDP and Education metrics. Of those 189 countries, St. Kitts and Nevis (KNA) was identified as the 13th lowest ranked by USD. It also becomes clear after anlayzing the data that "High income: OECD" countries rank higher in GDP than "High income: nonOECD" countries and that 75% of all "High income: OECD" country GDP data exceed at least 75% of each other income group's GDP values. Finally, combining these two data sets also allows for GDP ranking quantiles to be compared to income group classifications. By this means, it is discovered that five countries from the "Lower middle income" group also rank among the top thirty-eight countries with highest GDP. Of course, even with these newly gained insights, there is still much more left to discover from GDP and education data.
